---
title: "**Part 2**: Explore spatial resolutions with AIRR + SpaceRanger"
author: 
- name: Jonas Schuck
  affiliation: 
  - Goethe University, University Hospital Frankfurt, Neurological Institute (Edinger Institute), Frankfurt/Main
  - University Cancer Center, Frankfurt/Main
  - Frankfurt Cancer Institute, Frankfurt/Main
  email: schuck@med.uni-frankfurt.de
- name: Katharina Imkeller
  affiliation: 
  - Goethe University, University Hospital Frankfurt, Neurological Institute (Edinger Institute), Frankfurt/Main
  - University Cancer Center, Frankfurt/Main
  - Frankfurt Cancer Institute, Frankfurt/Main
  email: imkeller@med.uni-frankfurt.de
output: 
    github_document:
        html_preview: false
        df_print: kable
date: "2026-08-05"
#output: html_document
---

```{r, include=FALSE}
options(knitr.table.format = "html")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "../../images_design/images/spAIRR_vignette1/p2/",
  dpi = 150
)
```

This vignette provides a starting point for analysing spatial AIRR data
generated with LongAIRR. Using a **Visium HD 3’** dataset, we
demonstrate how to:

- load and inspect a LongAIRR AIRR rearrangement table
- summarize the receptor repertoire and assign IGH clones using SCOPer
- visualize clone distributions using LongAIRR spatial coordinates
- combine LongAIRR output with matching Space Ranger results and
- import the annotated AIRR table into scRepertoire

This is **Part 2** of the vignette. Refer to the [**Vignette 1 Overview**](./vignette1_overview.md)
or move to [**Part 1**](./longairr_vignette1_part1.md) and [**Part 3**](./longairr_vignette1_part3.md) directly.

---


```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)
library(readr)
library(scoper)
library(scRepertoire)
library(shazam)
library(stringr)
library(tibble)
library(cowplot)
```

------------------------------------------------------------------------

The analysis shown in the previous **Part 1** explores first steps of
downstream AIRR analysis using only AIRR data generated by LongAIRR.

When Visium HD 3′ datasets are processed with `LongAIRR-Whitelist` to
generate a spatial barcode whitelist, the matching Space Ranger run
provides the registered tissue image and the mapping between native 2 µm
spatial barcodes and larger visualization bins.

------------------------------------------------------------------------

This part uses the `igh_clones` dataframe from [**Part 1**](./longairr_vignette1_part1.md):

```{r}
# inspect clone df
igh_clones %>%
  select(sequence_id, v_call, j_call, isotype_sub, junction, clone_id, spbcid, x, y) %>%
  head(8)
```


## Provide SpaceRanger paths

```{r}
SPACERANGER_OUTS_DIR <- file.path("/path/to/longairr_results/runXYZ/spaceranger/outs")
SPACERANGER_IMAGE_DIR <- file.path(SPACERANGER_OUTS_DIR, "spatial")

SPACERANGER_2UM_DIR <-  file.path(SPACERANGER_OUTS_DIR, "binned_outputs", "square_002um", "spatial")
SPACERANGER_16UM_DIR <- file.path(SPACERANGER_OUTS_DIR, "binned_outputs", "square_016um", "spatial")

BARCODE_MAPPING_FILE <- file.path(SPACERANGER_OUTS_DIR, "barcode_mappings.parquet")

TISSUE_IMAGE_RESOLUTION <- "hires" # "lowres" or "hires"
CROP_PADDING_FRACTION <- 0.3 #used in this example to crop the tissue image

tissue_image_file <- file.path(SPACERANGER_IMAGE_DIR, paste0("tissue_", TISSUE_IMAGE_RESOLUTION, "_image.png"))
scalefactors_file <- file.path(SPACERANGER_16UM_DIR,"scalefactors_json.json")
```

## Map native 2 µm barcodes to 16 µm bins and segmented cells

Space Ranger stores the correspondence between bin resolutions in
`outs/barcode_mappings.parquet`. Only mappings for spatial barcodes observed
by LongAIRR are loaded into memory.

The `-1` suffix used by 10x Genomics identifies the library and is removed
before joining the different tables.

```{r}
remove_10x_suffix <- function(x) {
  sub("-1$", "", as.character(x))
}

observed_2um_ids <- igh_clones %>%
  transmute(spbc_2um = remove_10x_suffix(spbcid)) %>%
  distinct() %>%
  pull(spbc_2um)

mapping_query_ids <- unique(c(observed_2um_ids, paste0(observed_2um_ids, "-1")))

mapping_dataset <- arrow::open_dataset(BARCODE_MAPPING_FILE, format = "parquet")

barcode_mapping <- mapping_dataset %>%
  filter(square_002um %in% mapping_query_ids) %>%
  select(square_002um, square_016um, cell_id, in_cell) %>%
  collect() %>%
  transmute(spbc_2um = remove_10x_suffix(square_002um),
            spbc_16um = remove_10x_suffix(square_016um),
            cell_id = if_else(
            in_cell & cell_id != "cellid_000000000",
            remove_10x_suffix(cell_id),
            NA_character_))

igh_hd <- igh_clones %>%
  mutate(spbc_2um = remove_10x_suffix(spbcid)) %>%
  left_join(barcode_mapping, by = "spbc_2um")

# inspect annotated bins
igh_hd %>%
  select(sequence_id, v_call, j_call, isotype_sub, junction, clone_id, spbcid, x, y, spbc_2um, spbc_16um, cell_id) %>%
  head(8)
```

QC: Check if all 2 µm bins got mapped correctly

```{r}
mapping_summary <- tibble::tibble(
  metric = c("LongAIRR 2 µm positions", "Positions mapped to 16 µm", "Positions mapped to segmented cells", "Segmented cells containing LongAIRR molecules"),
  value = c(n_distinct(igh_hd$spbc_2um, na.rm = TRUE), n_distinct(igh_hd$spbc_2um[!is.na(igh_hd$spbc_16um)]),
            n_distinct(igh_hd$spbc_2um[!is.na(igh_hd$cell_id)]), n_distinct(igh_hd$cell_id, na.rm = TRUE)))

mapping_summary
```

## Summarize clone distributions at all spatial resolutions

At 2 µm, each summary refers to a native LongAIRR spatial barcode. At 16 µm, and
for the segmented cell annotation, all native squares assigned to the same Space Ranger bin are combined before
clone richness and molecule counts are calculated.

```{r}
top_clone_ids <- c("3")

richness_2um <- igh_hd %>%
  group_by(spbc_2um) %>%
  summarise(n_clones = n_distinct(clone_id), n_molecules = n(), .groups = "drop")

richness_16um <- igh_hd %>%
  group_by(spbc_16um) %>%
  summarise(n_clones = n_distinct(clone_id), n_molecules = n(), .groups = "drop")

richness_cell <- igh_hd %>%
  group_by(cell_id) %>%
  summarise(n_clones = n_distinct(clone_id), n_molecules = n(), .groups = "drop")

top_clones_2um <- igh_hd %>%
  filter(clone_id %in% top_clone_ids) %>%
  count(clone_id, spbc_2um, name = "n_molecules") %>%
  mutate(clone_id = factor(clone_id, levels = top_clone_ids))

top_clones_16um <- igh_hd %>%
  filter(clone_id %in% top_clone_ids) %>%
  count(clone_id, spbc_16um, name = "n_molecules") %>%
  mutate(clone_id = factor(clone_id, levels = top_clone_ids))

top_clones_cell <- igh_hd %>%
  filter(clone_id %in% top_clone_ids) %>%
  count(clone_id, cell_id, name = "n_molecules") %>%
  mutate(clone_id = factor(clone_id, levels = top_clone_ids))
```


## Add tissue coordinates

The tissue image and its scale factor are read from the Space Ranger output.
The native and aggregated summaries are then joined to the corresponding
`tissue_positions.parquet` files.

```{r}
tissue_image_file <- file.path(SPACERANGER_IMAGE_DIR, paste0("tissue_",TISSUE_IMAGE_RESOLUTION,"_image.png"))
scalefactors_file <- file.path(SPACERANGER_16UM_DIR,"scalefactors_json.json")

tissue_image <- png::readPNG(tissue_image_file)

# Rotate the tissue image 90 degrees clockwise to match SpaceRanger coordinates
tissue_image <- aperm(tissue_image[dim(tissue_image)[1]:1, , , drop = FALSE], c(2, 1, 3))

# Dimensions change after rotation.
image_height <- dim(tissue_image)[1]
image_width <- dim(tissue_image)[2]

scalefactors <- jsonlite::fromJSON(scalefactors_file)
scale_name <- paste0("tissue_",TISSUE_IMAGE_RESOLUTION,"_scalef")

image_scale <- scalefactors[[scale_name]]
```

The native 2 µm position table contains millions of rows. Therefore, only
positions observed in the LongAIRR output are collected into R.

```{r}
#observed 2um barcodes
position_query_ids_2um <- unique(c(richness_2um$spbc_2um, paste0(richness_2um$spbc_2um, "-1")))

positions_2um <- arrow::open_dataset(file.path(SPACERANGER_2UM_DIR,"tissue_positions.parquet"), format = "parquet") %>%
  filter(barcode %in% position_query_ids_2um) %>%
  select(barcode, pxl_row_in_fullres, pxl_col_in_fullres) %>%
  collect() %>%
  mutate(barcode = remove_10x_suffix(barcode),
         image_x = pxl_row_in_fullres * image_scale,
         image_y = pxl_col_in_fullres * image_scale)
```

The 16 µm position table is considerably smaller. All positions are retained
because they are also used below to determine the capture-area boundaries.

```{r}
positions_16um <- arrow::open_dataset(file.path(SPACERANGER_16UM_DIR,"tissue_positions.parquet"),format = "parquet") %>%
  select(barcode, pxl_row_in_fullres, pxl_col_in_fullres) %>%
  collect() %>%
  mutate(barcode = remove_10x_suffix(barcode),
         image_x = pxl_row_in_fullres * image_scale,
         image_y = pxl_col_in_fullres * image_scale)
```

Observed cells mapping

```{r }
observed_cell_ids <- igh_hd %>%
  distinct(cell_id) %>%
  pull(cell_id)

cell_query_ids <- unique(c(observed_cell_ids, paste0(observed_cell_ids, "-1")))

cell_bin_mapping <- mapping_dataset %>%
  filter(in_cell, cell_id %in% cell_query_ids) %>%
  select(square_002um, cell_id) %>%
  collect() %>%
  transmute(spbc_2um = remove_10x_suffix(square_002um),
            cell_id = remove_10x_suffix(cell_id))

cell_bin_query_ids <- unique(c(cell_bin_mapping$spbc_2um, paste0(cell_bin_mapping$spbc_2um, "-1")))

cell_positions <- arrow::open_dataset(file.path(SPACERANGER_2UM_DIR, "tissue_positions.parquet"),format = "parquet") %>%
  filter(barcode %in% cell_bin_query_ids) %>%
  select(barcode, pxl_row_in_fullres, pxl_col_in_fullres) %>%
  collect() %>%
  transmute(spbc_2um = remove_10x_suffix(barcode), 
            image_x = pxl_row_in_fullres * image_scale,
            image_y = pxl_col_in_fullres * image_scale) %>%
  inner_join(cell_bin_mapping, by = "spbc_2um") %>%
  group_by(cell_id) %>%
  summarise(image_x = mean(image_x), image_y = mean(image_y), .groups = "drop")
```


Finally, the clone summaries are joined to the registered image coordinates.

```{r}
richness_tissue_2um <- richness_2um %>%
  left_join(positions_2um, by = c("spbc_2um" = "barcode"))

richness_tissue_16um <- richness_16um %>%
  left_join(positions_16um, by = c("spbc_16um" = "barcode"))

richness_tissue_cell <- richness_cell %>%
  left_join(cell_positions, by = "cell_id")

top_clones_tissue_2um <- top_clones_2um %>%
  left_join(positions_2um, by = c("spbc_2um" = "barcode")) %>%
  filter(clone_id == "3")

top_clones_tissue_16um <- top_clones_16um %>%
  left_join(positions_16um, by = c("spbc_16um" = "barcode")) %>%
  filter(clone_id == "3")

top_clones_tissue_cell <- top_clones_cell %>%
  left_join(cell_positions, by = "cell_id") %>%
  filter(clone_id == "3")

#QC check number of bins per resolution
position_join_summary <- tibble::tibble(
  resolution = c("Native 2 µm", "Aggregated 16 µm", "Segmented cells"),
  n_spbcs = c(
    nrow(richness_tissue_2um),
    nrow(richness_tissue_16um),
    nrow(richness_tissue_cell)))

position_join_summary
```


## Spatial distribution comparison between resolutions

The displayed image is cropped to the barcoded capture area. Here we use the H&E image.
It is also possible to use the tissue image with the Visium fiducial frame and crop it accordingly.
Increase `CROP_PADDING_FRACTION` if more of the frame should remain visible.

```{r, fig.width=15, fig.height=13}
# Defining image boundaries, also zoom-in boundaries on the tissue layer

capture_bounds <- positions_16um %>%
  summarise(xmin = min(image_x, na.rm = TRUE),
            xmax = max(image_x, na.rm = TRUE),
            ymin = min(image_y, na.rm = TRUE),
            ymax = max(image_y, na.rm = TRUE))

x_padding <- (capture_bounds$xmax - capture_bounds$xmin) * CROP_PADDING_FRACTION
y_padding <-(capture_bounds$ymax - capture_bounds$ymin) * CROP_PADDING_FRACTION

crop_x_limits <- c(max(0, capture_bounds$xmin - x_padding), min(image_width, capture_bounds$xmax + x_padding))
crop_y_limits <- c(min(image_height, capture_bounds$ymax + y_padding), max(0, capture_bounds$ymin - y_padding))

# Select one clone and a region for closer inspection.
ZOOM_CLONE <- as.character(top_clone_ids[1])  # clone 3 in this example

x_span <- diff(crop_x_limits)
y_span <- abs(diff(crop_y_limits))

# Bottom-left portion of the capture area (clone rich area)
# define how small the zoomed-box should be with the scaling factor (0.25 here)
zoom_x_limits <- c(crop_x_limits[1], crop_x_limits[1] + 0.4 * x_span)
zoom_y_limits <- c(max(crop_y_limits), max(crop_y_limits) - 0.4 * y_span)

# Keep clone colours consistent between the complete and zoomed plots.
clone_colours <- scales::hue_pal()(length(top_clone_ids))
names(clone_colours) <- as.character(top_clone_ids)

zoom_box <- tibble::tibble(clone_id = factor(ZOOM_CLONE, levels = top_clone_ids),
                           xmin = min(zoom_x_limits), xmax = max(zoom_x_limits),
                           ymin = min(zoom_y_limits), ymax = max(zoom_y_limits))

#cropped tissue layer
tissue_background <- annotation_raster(tissue_image,
  xmin = 0, xmax = image_width,
  ymin = image_height, ymax = 0)

richness_limits <- range(c(richness_tissue_2um$n_clones, richness_tissue_16um$n_clones, richness_tissue_cell$n_clones), na.rm = TRUE)
```

### Clonal richness for 2 µm, 16 µm and segmented cells

```{r}
p_richness_2um <- ggplot() +
  tissue_background +
  geom_point(data = richness_tissue_2um %>%
                    filter(!is.na(image_x), !is.na(image_y)),
            aes(x = image_x, y = image_y, fill = n_clones),
            shape = 21, color = "black", stroke = 0.4, size = 1.5, alpha = 0.9) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = crop_x_limits, ylim = crop_y_limits, expand = FALSE) +
  labs(title = "Native 2 µm bins") +
  theme_void(base_size = 9)

p_richness_16um <- ggplot() +
  tissue_background +
  geom_point(data = richness_tissue_16um %>%
                    filter(!is.na(image_x), !is.na(image_y)),
    aes(x = image_x, y = image_y, fill = n_clones),
    shape = 21, color = "black", stroke = 0.4, size = 1.5, alpha = 0.95) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = crop_x_limits, ylim = crop_y_limits, expand = FALSE) +
  labs(title = "Aggregated 16 µm bins") +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5))

p_richness_cell <- ggplot() +
  tissue_background +
  geom_point(data = richness_tissue_cell %>%
                    filter(!is.na(image_x), !is.na(image_y)),
    aes(x = image_x, y = image_y, fill = n_clones),
    shape = 21, color = "black", stroke = 0.4, size = 1.5, alpha = 0.95) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = crop_x_limits, ylim = crop_y_limits, expand = FALSE) +
  labs(title = "Segmented cells") +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5))

p_richness_bins <- (p_richness_2um | p_richness_16um | p_richness_cell) +
                   patchwork::plot_annotation(title = "Clonal richness")

p_richness_bins
```


### Spatial distribution of top clone for 2 µm, 16 µm and segmented cells

```{r}
p_top_2um <- ggplot() + tissue_background +
  geom_point(data = top_clones_tissue_2um %>%
             filter(!is.na(image_x), !is.na(image_y)),
    aes(x = image_x, y = image_y, fill = clone_id, size = n_molecules),
    shape = 21,color = "black", stroke = 0.4, alpha = 0.9) +
  facet_wrap(~clone_id, nrow = 1) +
  scale_size_continuous(trans = "sqrt", name = "Molecules", range = c(1.5, 4)) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = crop_x_limits, ylim = crop_y_limits, expand = FALSE) +
  scale_fill_manual(values = clone_colours, drop = FALSE) +
  labs(title = paste0("Native 2 µm")) +
  guides(fill = "none") +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5), strip.text = element_text(size = 9))

p_top_16um <- ggplot() + tissue_background +
  geom_point(data = top_clones_tissue_16um %>%
             filter(!is.na(image_x), !is.na(image_y)),
    aes(x = image_x, y = image_y, fill = clone_id, size = n_molecules),
    shape = 21, color = "black", stroke = 0.4, alpha = 0.95) +
  facet_wrap(~clone_id, nrow = 1) +
  scale_size_continuous(trans = "sqrt", name = "Molecules", range = c(1.5, 4)) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = crop_x_limits, ylim = crop_y_limits, expand = FALSE) +
  scale_fill_manual(values = clone_colours, drop = FALSE) +
  labs(title = paste0("Aggregated 16 µm")) +
  guides(fill = "none") +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5), strip.text = element_text(size = 9)) 

p_top_cell <- ggplot() + tissue_background +
  geom_point(data = top_clones_tissue_cell %>%
             filter(!is.na(image_x), !is.na(image_y)),
    aes(x = image_x, y = image_y, fill = clone_id, size = n_molecules),
    shape = 21, color = "black", stroke = 0.4, alpha = 0.95) +
  facet_wrap(~clone_id, nrow = 1) +
  scale_size_continuous(trans = "sqrt", name = "Molecules", range = c(1.5, 4)) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = crop_x_limits, ylim = crop_y_limits, expand = FALSE) +
  scale_fill_manual(values = clone_colours, drop = FALSE) +
  labs(title = paste0("Segmented cells")) +
  guides(fill = "none") +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5), strip.text = element_text(size = 9)) 

p_top_clones_bins <- (p_top_2um | p_top_16um | p_top_cell) +
                    patchwork::plot_annotation(title = "Spatial distribution of top clone")

p_top_clones_bins
```


Figures for the zoomed-in, clone-rich area for 3 resolutions.
And complete figure panel

```{r, fig.width=9, fig.height=9, fig.align="center"}
p_zoom_2um <- ggplot() +
  tissue_background +
  geom_point(data = top_clones_tissue_2um %>%
      filter(as.character(clone_id) == ZOOM_CLONE),
    aes(x = image_x, y = image_y, fill = clone_id, size = n_molecules),
    shape = 21, color = "black", stroke = 0.4, alpha = 0.95) +
  scale_fill_manual(values = clone_colours, limits = names(clone_colours), drop = FALSE, guide = "none") +
  scale_size_continuous(trans="identity", name = "Molecules", range = c(1, 4)) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = zoom_x_limits, ylim = zoom_y_limits, expand = FALSE) +
  labs(title = paste0("Clone ", ZOOM_CLONE, ": native 2 µm")) +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5))

p_zoom_16um <- ggplot() +
  tissue_background +
  geom_point(data = top_clones_tissue_16um %>%
      filter(as.character(clone_id) == ZOOM_CLONE),
    aes(x = image_x, y = image_y, fill = clone_id, size = n_molecules),
    shape = 21, color = "black", stroke = 0.4, alpha = 0.95) +
  scale_fill_manual(values = clone_colours, limits = names(clone_colours), drop = FALSE, guide = "none") +
  scale_size_continuous(trans = "identity", name = "Molecules", range = c(1, 4)) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = zoom_x_limits, ylim = zoom_y_limits, expand = FALSE) +
  labs(title = paste0("Clone ", ZOOM_CLONE,": aggregated 16 µm")) +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5))

p_zoom_cell <- ggplot() +
  tissue_background +
  geom_point(data = top_clones_tissue_cell %>%
      filter(as.character(clone_id) == ZOOM_CLONE),
    aes(x = image_x, y = image_y, fill = clone_id, size = n_molecules),
    shape = 21, color = "black", stroke = 0.4, alpha = 0.95) +
  scale_fill_manual(values = clone_colours, limits = names(clone_colours), drop = FALSE, guide = "none") +
  scale_size_continuous(trans = "identity", name = "Molecules", range = c(1, 4)) +
  scale_y_reverse(limits = c(image_height, 0), expand = expansion(0)) +
  coord_fixed(xlim = zoom_x_limits, ylim = zoom_y_limits, expand = FALSE) +
  labs(title = paste0("Clone ", ZOOM_CLONE,": segmented cells")) +
  theme_void(base_size = 9) +
  theme(plot.title = element_text(hjust = 0.5))

p_top_clones_bins_zoomed <- (p_zoom_2um / p_zoom_16um / p_zoom_cell) +
                            patchwork::plot_annotation(title = "zoom-in (bottom left)")

p_top_clone_combined <- plot_grid(p_top_clones_bins, p_top_clones_bins_zoomed, ncol=2)
p_top_clone_combined
```


The selected clone is visualized on the registered H&E image at native 2 µm resolution, after aggregation into 16 µm bins, and after assignment to Space Ranger-segmented cells. Point size represents the number of AIRR molecules detected at each position. The zoomed panels highlight a receptor-dense region and illustrate how spatial aggregation affects visualization while preserving the overall localization pattern.

---

**Part 3** will show how to read in the generated `igh_clones` dataframe into
**`scRepertoire`**. This serves as interoperability-check for using additional established AIRR analysis frameworks
for downstream analysis of spatial AIRR datasets generated using LongAIRR.

---

# Additional resources

- [LongAIRR GitHub repository](https://github.com/AGImkeller/LongAIRR)  
  Source code, installation instructions, example workflows and issue tracker.

- [LongAIRR-Whitelist GitHub repository](https://github.com/AGImkeller/LongAIRR_whitelist)  
  Companion workflow for generating sample-specific Visium HD 3′ spatial-barcode indices.  
  *This link will become available once the repository is public.*

- [AIRR Community Data Standards](https://docs.airr-community.org/en/latest/datarep/overview.html)  
  Standards for representing and exchanging adaptive immune receptor repertoire data.

- [AIRR Rearrangement schema](https://docs.airr-community.org/en/latest/datarep/rearrangements.html)  
  Definitions of the standardized fields used in AIRR rearrangement tables.

- [Immcantation documentation](https://immcantation.readthedocs.io/en/stable/)  
  Framework for AIRR-seq processing and downstream repertoire analysis.

- [SCOPer documentation](https://scoper.readthedocs.io/)  
  Methods for assigning B-cell clonal relationships from AIRR-formatted data.

- [scRepertoire on Bioconductor](https://bioconductor.org/packages/scRepertoire/)  
  Installation, reference manual and vignettes for single-cell immune-receptor analysis.

- [scRepertoire GitHub repository](https://github.com/BorchLab/scRepertoire)  
  Source code, examples and issue tracker.


---

# Session Information

```{r}
sessionInfo()
```


