---
title: "**Part 3**: Interoperability with `scRepertoire`"
author: 
- name: Jonas Schuck
  affiliation: 
  - Goethe University, University Hospital Frankfurt, Neurological Institute (Edinger Institute), Frankfurt/Main
  - University Cancer Center, Frankfurt/Main
  - Frankfurt Cancer Institute, Frankfurt/Main
  email: schuck@med.uni-frankfurt.de
- name: Katharina Imkeller
  affiliation: 
  - Goethe University, University Hospital Frankfurt, Neurological Institute (Edinger Institute), Frankfurt/Main
  - University Cancer Center, Frankfurt/Main
  - Frankfurt Cancer Institute, Frankfurt/Main
  email: imkeller@med.uni-frankfurt.de
output: 
    github_document:
        html_preview: false
        df_print: kable
date: "2026-08-05"
#output: html_document
---

```{r, include=FALSE}
options(knitr.table.format = "html")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "../../images_design/images/spAIRR_vignette1/p3/",
  dpi = 150
)
```

This vignette provides a starting point for analysing spatial AIRR data
generated with LongAIRR. Using a **Visium HD 3’** dataset, we
demonstrate how to:

- load and inspect a LongAIRR AIRR rearrangement table
- summarize the receptor repertoire and assign IGH clones using SCOPer
- visualize clone distributions using LongAIRR spatial coordinates
- combine LongAIRR output with matching Space Ranger results and
- import the annotated AIRR table into scRepertoire

This is **Part 3** of the vignette. Refer to the [**Vignette 1 Overview**](./vignette1_overview.md)
or move to [**Part 1**](./longairr_vignette1_part1.md) and [**Part 2**](./longairr_vignette1_part2.md) directly.

---

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)
library(readr)
library(scoper)
library(scRepertoire)
library(shazam)
library(stringr)
library(tibble)
library(cowplot)
```

------------------------------------------------------------------------

scRepertoire is designed around single-cell barcodes. For this
interoperability example, every LongAIRR **UMI-SPBCID molecule group**
is assigned a unique technical `cell_id`. This preserves one
reconstructed molecule as one input unit, but it does **not** convert a
spatial position or molecule group into a biologically resolved cell.
Resulting abundances must therefore be interpreted as molecule counts
rather than cell counts.

------------------------------------------------------------------------

This part uses the `igh_clones` dataframe from [**Part 1**](./longairr_vignette1_part1.md):

```{r}
# inspect clone df
igh_clones %>%
  select(sequence_id, v_call, j_call, isotype_sub, junction, clone_id, spbcid, x, y) %>%
  head(8)
```


# Using scRepertoire for spAIRR repertoire exploration

scRepertoire is designed around single-cell barcodes. For this interoperability
example, every LongAIRR **UMI-SPBCID molecule group** is assigned a unique technical
`cell_id`. This preserves one reconstructed molecule as one input unit, but it
does **not** convert a spatial position or molecule group into a biologically
resolved cell. Resulting abundances must therefore be interpreted as molecule
counts rather than cell counts.

## Import the annotated AIRR table into scRepertoire

```{r}
SAMPLE_LABEL <- "spAIRR_HD"

sc_airr_input <- igh_clones %>%
  mutate(cell_id = as.character(group))

group_clone_conflicts <- sc_airr_input %>%
  distinct(cell_id, clone_id) %>%
  count(cell_id, name = "n_clone_ids") %>%
  filter(n_clone_ids > 1)

sc_airr_list <- list(as.data.frame(sc_airr_input))
names(sc_airr_list) <- SAMPLE_LABEL

sc_contig_list <- scRepertoire::loadContigs(input = sc_airr_list, format = "AIRR")
sc_bcr <- scRepertoire::combineBCR(input.data = sc_contig_list, samples = SAMPLE_LABEL, chain = "IGH")
```

`combineBCR()` reshapes the AIRR input and may prefix its technical barcode (append the sample-name to the group-ID, e.g. `spAIRR_HD_AATACTAGTs_002um_00717_00138-1`). 
The earlier annotated **clone_id** is omitted during this process.
We therefore join it back using the original molecule-group identifier (umi-spbcid)

```{r}
#lookup (umi_spbcid) to cloneid
sc_clone_lookup <- sc_airr_input %>%
  transmute(group_for_join = as.character(cell_id),
            clone_id = as.character(clone_id)) %>%
  distinct()

# merges back the clone_id to the corresponding group ID using lookup // required to restore SCOPer clone IDs 
sc_bcr <- purrr::imap(sc_bcr,
  function(df, sample_name) {
    df %>%
      select(-any_of("clone_id")) %>%
      mutate(group_for_join = stringr::str_remove(as.character(barcode), paste0("^", SAMPLE_LABEL, "_"))) %>%
      left_join(sc_clone_lookup, by = "group_for_join")
  }
)
```

scRepertoire provides several functions for repertoire-exploration that serve as a starting point for every downstream analysis.
Below are two examples:

```{r}
#used functions expect previous use of combineBCR or combineTCR
p_sc_prop <- scRepertoire::clonalProportion(input.data = sc_bcr,
                                            clone.call = "clone_id",
                                            clonal.split = c(1, 3, length(unique(sc_bcr$spAIRR_HD$clone_id))),
                                            chain = "IGH")

# IGHv genes overview
p_sc_genes <- scRepertoire::vizGenes(sc_bcr, x.axis = "IGHV", y.axis = NULL,
                                     plot = "barplot", summary.fun = "proportion") 

p_screp_combined <- plot_grid(p_sc_prop, p_sc_genes, ncol=2)
p_screp_combined
```

---

Move to the vignette overview to explore **Part 1** and **Part 2** of this vignette.

---

# Additional resources

- [LongAIRR GitHub repository](https://github.com/AGImkeller/LongAIRR)  
  Source code, installation instructions, example workflows and issue tracker.

- [LongAIRR-Whitelist GitHub repository](https://github.com/AGImkeller/LongAIRR_whitelist)  
  Companion workflow for generating sample-specific Visium HD 3′ spatial-barcode indices.  
  *This link will become available once the repository is public.*

- [AIRR Community Data Standards](https://docs.airr-community.org/en/latest/datarep/overview.html)  
  Standards for representing and exchanging adaptive immune receptor repertoire data.

- [AIRR Rearrangement schema](https://docs.airr-community.org/en/latest/datarep/rearrangements.html)  
  Definitions of the standardized fields used in AIRR rearrangement tables.

- [Immcantation documentation](https://immcantation.readthedocs.io/en/stable/)  
  Framework for AIRR-seq processing and downstream repertoire analysis.

- [SCOPer documentation](https://scoper.readthedocs.io/)  
  Methods for assigning B-cell clonal relationships from AIRR-formatted data.

- [scRepertoire on Bioconductor](https://bioconductor.org/packages/scRepertoire/)  
  Installation, reference manual and vignettes for single-cell immune-receptor analysis.

- [scRepertoire GitHub repository](https://github.com/BorchLab/scRepertoire)  
  Source code, examples and issue tracker.


---

# Session Information

```{r}
sessionInfo()
```







