---
title: "**Part 1**: Start spatial AIRR analysis with LongAIRR output"
author: 
- name: Jonas Schuck
  affiliation: 
  - Goethe University, University Hospital Frankfurt, Neurological Institute (Edinger Institute), Frankfurt/Main
  - University Cancer Center, Frankfurt/Main
  - Frankfurt Cancer Institute, Frankfurt/Main
  email: schuck@med.uni-frankfurt.de
- name: Katharina Imkeller
  affiliation: 
  - Goethe University, University Hospital Frankfurt, Neurological Institute (Edinger Institute), Frankfurt/Main
  - University Cancer Center, Frankfurt/Main
  - Frankfurt Cancer Institute, Frankfurt/Main
  email: imkeller@med.uni-frankfurt.de
output: 
    github_document:
        html_preview: false
        df_print: kable
date: "2026-08-05"
#output: html_document
---

```{r, include=FALSE}
options(knitr.table.format = "html")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "../../images_design/images/spAIRR_vignette1/p1/",
  dpi = 150
)
```

This vignette provides a starting point for analysing spatial AIRR data
generated with LongAIRR. Using a **Visium HD 3’** dataset, we
demonstrate how to:

- load and inspect a LongAIRR AIRR rearrangement table
- summarize the receptor repertoire and assign IGH clones using SCOPer
- visualize clone distributions using LongAIRR spatial coordinates
- combine LongAIRR output with matching Space Ranger results and
- import the annotated AIRR table into scRepertoire

This is **Part 1** of the vignette. Refer to the [**Vignette 1 Overview**](./vignette1_overview.md)
or move to [**Part 2**](./longairr_vignette1_part2.md) and [**Part 3**](./longairr_vignette1_part3.md) directly.

**Part 2** and **Part 3** will reuse the main object generated in this
chapter.

---


```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)
library(readr)
library(scoper)
library(scRepertoire)
library(shazam)
library(stringr)
library(tibble)
library(cowplot)
```

------------------------------------------------------------------------

This first part requires only the AIRR rearrangement table produced by
LongAIRR. No Space Ranger files or gene-expression data are used.

## Load and inspect the LongAIRR output

Refer to the (spatial) AIRR chapter for more information about the expected
columns in spatial AIRR tables.

```{r}
SAMPLE_LABEL <- "spAIRR_HD"

AIRR_FILE <-file.path("/path/to/longairr_results/runXYZ/airr/ig_p_parse-select.tsv")

airr_raw <- readr::read_tsv(AIRR_FILE, show_col_types = FALSE, progress = FALSE)
```

```{r}
#perform downstream filters
airr_filtered <- airr_raw %>% 
                filter(!is.na(c_call) & str_sub(c_call, 1, 3) == locus & !grepl("TTTTTTTTTTT", umi) & !grepl("N", junction)) %>%
                filter(consensus_count > 1)

# retrieve sub-isotype from c_call annotation
airr <- airr_filtered %>%
  mutate(isotype_sub = stringr::str_extract(as.character(c_call), "IGH[A-Z][A-Z0-9]*"))

table(airr$isotype_sub)
```

The following summaries provide a quick overview of the selected input and
identify fields that would prevent clone assignment or spatial visualization.

```{r}
dataset_summary <- tibble::tribble(
  ~metric, ~value,
  "Productive rows", sum(airr$productive),
  "Productive IGH rows", sum(airr$productive & airr$locus == "IGH", na.rm = TRUE),
  "Spatial positions [spbcid]", dplyr::n_distinct(airr$spbcid, na.rm = TRUE),
  "Molecule groups [group]", dplyr::n_distinct(airr$group, na.rm = TRUE))

dataset_summary
```

---

## Repertoire overview

Before assigning clones, simple summaries can be generated directly from the
AIRR table. Here, each row represents one reconstructed LongAIRR molecule.

```{r}
locus_counts <- airr %>%
  count(locus, name = "n_chains")

isotype_counts <- airr %>%
  filter(locus == "IGH") %>%
  count(isotype_sub, name = "n_chains") %>%
  arrange(desc(n_chains))
```

Complete figure:

```{r, fig.height=3}
p_locus <- ggplot(locus_counts, aes(x = locus, y = n_chains, fill = locus)) +
  geom_col(width = 0.75, show.legend = FALSE) +
  labs(title = "BCR chain composition", x = "Ig Locus", y = "Count") +
  theme_bw(base_size = 9)

p_isotype <- ggplot(isotype_counts, aes(x = isotype_sub, y = n_chains, fill = isotype_sub)) +
  geom_col(width = 0.75, show.legend = FALSE) +
  labs(title = "IGH isotype composition", x = "Ig Isotype", y = "Count") +
  theme_bw(base_size = 9)

p_repertoire_overview <- p_locus + p_isotype
p_repertoire_overview
```


---

## Assign IGH clones with SCOPer

SCOPer clone assignment in this vignette focuses on **productive IGH sequences**.
Spatial coordinates are retained as metadata but are not used to define the
clones.

SCOPer applies hierarchical clustering using the selected threshold and adds a
new `clone_id` column. We choose `0.1` here. Clone identifiers are specific to the selected
dataset and clone-assignment parameters.

We further generate a figure giving an overview about the clone-size distribution.

```{r}
airr_igh <- airr %>%
  filter(locus == "IGH")

scoper_result <- scoper::hierarchicalClones(db = airr_igh, threshold = 0.1, summarize_clones = TRUE)
igh_clones <- as.data.frame(scoper_result)

# inspect clone df
igh_clones %>%
  select(sequence_id, v_call, j_call, isotype_sub, junction, clone_id, spbcid, x, y) %>%
  head(8)
```

```{r}
#rank clones by size
clone_summary <- igh_clones %>%
  group_by(clone_id) %>%
  summarise(n_molecules = n(),
            n_spbcs = n_distinct(spbcid),
            sum_consensus_count = sum(consensus_count, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(n_molecules), clone_id) %>%
  mutate(clone_rank = row_number())

#inspect clone sizes
clone_summary %>%
  head(10)
```

```{r, fig.width=6.5, fig.height=4.5}
p_clone_rank <- ggplot(clone_summary,
  aes(x = clone_rank, y = n_molecules)) +
  geom_line(linewidth = 0.4, color = "grey50") +
  geom_point(size = 1.8) +
  scale_y_log10() +
  labs(title = "Clone-size distribution", x = "Clone rank", y = "Clone size (log10)") +
  theme_bw(base_size = 9)

p_clone_rank
```

---

## Spatial distribution of annotated IgH clones

To explore the clone distribution in their tissue context, we generate a figure
showcasing clone richness per spatial position and the distribution of the
most expanded clone

For the background layer we take the raw, unfiltered AIRR input to resemble the
overall tissue structure.

```{r}
#background layer of all captured spots
spatial_background <- airr_raw %>%
  distinct(spbcid, x, y)
```

Clonal richness:

```{r}
spatial_clone_summary <- igh_clones %>%
  group_by(spbcid, x, y) %>%
  summarise(n_clones = n_distinct(clone_id),
            n_molecules = n(), .groups = "drop")

p_richness <- ggplot() +
  geom_point(data = spatial_background, aes(x = y, y = -x), color = "grey80", size = 1.2) +     #background layer
  geom_point(data = spatial_clone_summary, aes(x = y, y = -x, color = n_clones), size = 1.2) +
  coord_equal() +
  labs(title = "IGH clone richness") +
  theme_void(base_size = 9)
```

Top clones:

```{r, fig.width=12, fig.height=5}
top_clone_ids <- clone_summary %>%
  slice_head(n = 2) %>%
  pull(clone_id)

top_clone_spatial <- igh_clones %>%
  filter(clone_id %in% top_clone_ids) %>%
  count(clone_id, spbcid, x, y, name = "n_molecules")
```

```{r, fig.width=12, fig.height=6}
p_top_clones <- ggplot() +
  geom_point(data = spatial_background, aes(x = y, y = -x), color = "grey80", size = 1.2) +
  geom_point(data = top_clone_spatial, aes(x = y, y = -x, color = clone_id, size = n_molecules), alpha = 0.9) +
  facet_wrap(~clone_id) +
  scale_size_continuous(name = "Molecules", range = c(1.2, 3.6)) +
  coord_equal() +
  labs(title = paste0("Spatial distribution of the top clones"),) +
  theme_bw(base_size = 9) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())
```

Complete figure:

```{r, fig.width=10, fig.height=6}
spatial_overview_complete <- plot_grid(p_richness, p_top_clones, ncol=2, rel_widths = c(0.5, 1))
spatial_overview_complete
```

Since the native spatial bin-size for Visium HD 3' datasets annotated with LongAIRR is 2 µm, accumulation of multiple clones within the same bin is rare.

We will explore how to assign the 16 µm spatial bin and segmented cell annotation provided by SpaceRanger in **Part 2**.

---

The next section of this vignette will show **how to combine spatial AIRR output from LongAIRR**
**with matching SpaceRanger output**. Specifically, how to aggregate the **native 2 µm resolution from Visium HD 3'** 
to the corresponding **16 µm** or **segmented cell bin-annotation** with ways of visualizing the spatial distribution
using the **H&E tissue images**.

---

# Additional resources

- [LongAIRR GitHub repository](https://github.com/AGImkeller/LongAIRR)  
  Source code, installation instructions, example workflows and issue tracker.

- [LongAIRR-Whitelist GitHub repository](https://github.com/AGImkeller/LongAIRR_whitelist)  
  Companion workflow for generating sample-specific Visium HD 3′ spatial-barcode indices.  
  *This link will become available once the repository is public.*

- [AIRR Community Data Standards](https://docs.airr-community.org/en/latest/datarep/overview.html)  
  Standards for representing and exchanging adaptive immune receptor repertoire data.

- [AIRR Rearrangement schema](https://docs.airr-community.org/en/latest/datarep/rearrangements.html)  
  Definitions of the standardized fields used in AIRR rearrangement tables.

- [Immcantation documentation](https://immcantation.readthedocs.io/en/stable/)  
  Framework for AIRR-seq processing and downstream repertoire analysis.

- [SCOPer documentation](https://scoper.readthedocs.io/)  
  Methods for assigning B-cell clonal relationships from AIRR-formatted data.

- [scRepertoire on Bioconductor](https://bioconductor.org/packages/scRepertoire/)  
  Installation, reference manual and vignettes for single-cell immune-receptor analysis.

- [scRepertoire GitHub repository](https://github.com/BorchLab/scRepertoire)  
  Source code, examples and issue tracker.


---

# Session Information

```{r}
sessionInfo()
```


